\documentclass[11pt]{article}
\usepackage[margin=0.8in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{hyperref}

\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{Y}{>{\raggedright\arraybackslash}X}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=blue,
}

\title{Reward Components}
\author{}
\date{April 5, 2026}

\begin{document}
\maketitle
\thispagestyle{empty}

\noindent
This table summarizes the reward logic used in the staged RL pipeline. Values below describe the raw reward behavior before phase-specific weighting.

\vspace{0.8em}

\small
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.25}

\begin{tabularx}{\linewidth}{L{2.4cm} L{3.0cm} Y L{4.1cm}}
\toprule
\textbf{Reward Type} & \textbf{What It Encourages} & \textbf{When It Gives High Reward} & \textbf{Simple Example} \\
\midrule

\textbf{Correctness}
& Get the right answer
& Exact normalized match with the ground truth, or the correct option letter for multiple-choice.
& Gold = 42, model = 42.0 $\rightarrow$ high reward \\

\textbf{Formatting}
& Follow the required structure
& Exactly one REASONING block and one SOLUTION block; malformed answers reduce the score.
& Clean reasoning + solution blocks $\rightarrow$ high reward \\

\textbf{Parseability}
& Make the solution machine-readable
& The solution can be parsed as a valid number or valid option letter.
& ``42'' $\rightarrow$ parseable; ``forty-two'' $\rightarrow$ low reward \\

\textbf{Finished}
& End the response cleanly
& The answer is complete and stays below the token limit.
& Properly closed answer $\rightarrow$ high reward; truncated at cap $\rightarrow$ low reward \\

\textbf{Tolerance}
& Be approximately correct on numeric tasks
& The answer is numerically close enough, but not already exact.
& Gold = 100, model = 99.999 $\rightarrow$ high reward \\

\textbf{Brevity}
& Stay concise
& The response is short relative to the allowed completion budget.
& 20 tokens $\rightarrow$ high reward; very long rambling answer $\rightarrow$ low reward \\

\bottomrule
\end{tabularx}

\vspace{0.8em}
\footnotesize
\noindent
\textbf{Note.} The required tags in code are exactly \texttt{<REASONING>} and \texttt{<SOLUTION>}. Tolerance reward is enabled only in later phases and does not add extra reward when the answer is already exact.

\end{document}
